|

### 2025-09-05 manuscript/paper.Rmd
ueahrsfdhdfgj

---
title: |
  | \large Assessing Ranking Fidelity of Competition Structures via Weighted Mutual Information
author: | 
  | \large Zach Culp \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  |
  | \large Josie Peterburs \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  |
  | \large Ryan McShane \vspace{-1.1mm}
  | \normalsize University of Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  |
  | \large Gregory J. Matthews \vspace{-1.1mm}
  |	\normalsize  \vspace{-1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Center for Data Science and Consulting \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`email`](mailto:ypu@something.edu) \vspace{-1mm}
  | \normalsize [`zculp@luc.edu`](mailto:zculp@luc.edu) \vspace{-1mm}
  | \normalsize [`jpeterburs@luc.edu`](mailto:jpeterburs@luc.edu) \vspace{-1mm}
  | \normalsize [`rmcshane@uchicago.edu`](mailto:rmcshane@uchicago.edu) \vspace{-1mm}
  | \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1mm}
abstract: | 
 | We evaluate multiple competitions structures to assess their effectiveness in accurately reproducing the true underlying ranking of participating teams. Using repeated simulations for each structure, we compare teams’ true performance ranks to their simulated outcomes. To quantify the accuracy of each format, we calculate weighted mutual information between true and simulated rankings across a large number of replicates. "Remarkable" weights are applied to look at only the top n teams/competitors in a given competition structure. This information-theoretic approach allows us to objectively compare formats and identify which structures most effectively preserve rank order. Results are visualized through comparative plots, providing clear insights into the trade-offs and strengths of each competition design.
 | 
  \vspace{2mm}
  | Keywords: tournament structures, mutual information
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
editor_options: 
  markdown: 
    wrap: 72
---

@@

Here, we propose a novel metric for evaluating different competition structures.

@@

Different competition structures have unique strengths and weaknesses, affecting how well they reflect the true rankings of the teams. The goal of a tournament is to find the best teams, but how can we quantify how well a tournament performs? In some instances, the tournament organizers only award the overall winner, but other times, the top three or ten teams are awarded. Because of this, organizers may prefer one structure over another based on their needs. Tournament organizers must also take into account factors like cost, timeliness, entertainment value, and fairness when choosing a tournament structure.

To evaluate a tournament’s effectiveness, we propose a numerical metric
that quantifies how accurately it orders teams based on their true
rankings. Tournament results can be viewed as a "message" attempting to
convey the true team rankings. However, various factors introduce noise,
leading to information loss. We use principles from information theory
to measure this information loss across multiple tournament simulations
to assess the reliability of different formats.

@@

Consider a set of $n$ teams indexed from $i = 1, 2, \cdots, n$ each with
an associated true strength parameter
$\boldsymbol{\theta} = (\theta_1, \theta_2, \ldots, \theta_n)$ such that
$(\theta_1 > \theta_2 > \ldots > \theta_n)$. Next, let
$r(\boldsymbol{\theta})$ be the indexes of $\boldsymbol{\theta}$ when
theta is sorted from largest to smallest so that
$r(\boldsymbol{\theta}) = \{1, 2, \ldots, n-1, n\}$. Next define
$T(\phi, s)$ as a tournament with schedule $\phi$ and seeding structure
$s$. $\phi$ contains all the information about the scheduling of teams,
which could be fully known prior to the tournament (i.e. round robin) or
determined as the tournament progresses (i.e. single elimination
tournament). We then let $\hat{r}(T(\phi, s),\boldsymbol{\theta})$ be an
$n$- dimensional vector valued random variable that gives the results of
a tournament as a vector of the indexes of the vector
$\boldsymbol{\theta}$. The index in the first position of the vector
$\hat{r}(T(\phi, s),\boldsymbol{\theta})$ indicates the index of the
team that finished first, the index in the second position of the vector
$\hat{r}(T(\phi, s),\boldsymbol{\theta})$ indicates the index of the
team that finished second, and so on.

@@

If one views the true ranking of the teams in a tournament as a message
to be sent to a receiver and the outcome of the tournament as a message
that is received, we can measure the "goodness" of a tournament in terms
of its ability to accurately transmit the true ranking. We can then use
concepts from information theory to assess the ability of a tournament
to correctly rank teams. Specifically, we start with the concept of
mutual information (CITE). For two random variables $X$ and $Y$ mutual
information is defined as:

@@

In order to alleviate this problem, we instead consider *weighted*
mutual information. We want to give more weight to permutations from
$\hat{r}$ that are "closer" to $r$. There are any number of reasonable
choices for this weighting function, and we choose to use squared error.
Specifically,

$$
w(r, \hat{r}) =
\begin{cases}
\frac{1}{\hat{r}'r}, & r \ne \hat{r} \\
1, & r = \hat{r}
\end{cases}
$$

In general, though any loss function $l(r, \hat{r})$ can be used:

@@

Note that one is 1/N!

And this does work because consistently getting the order incorrect is
perfect information.

However, because the number of teams that need to be accurately ordered
differs based on the tournament organizers, a second weighting function
was found to portray the importance, or remarkability, of each value in
$\hat{r}$. We call these "remarkable weights" ($m(r)$) to differentiate
from the other weighting function. "Remarkable weights" are set to allow
for a subjective belief of what determines a good tournament. For a
four-team tournament, if the user decides that only the winner is
important, then $m(r) = \{1,0,0,0\}$. However, if the organizers want an
accurate leaderboard with the winner as the most important, then a
potential $m(r)$ could be $\{3,1,1,0\}$. With the addition of the
weighting functions, the formula for mutual information is updated:

@@

Because this mutual information is unitless, the formula can be
standardized to be between 0 and 1 using the complement of Rajski's
distance: $$
1 - (1 - \frac{I(X,Y)}{\text{max\{}H(X), H(Y)\}})
$$

>>>>

---
title: |
  | \large Assessing Ranking Fidelity of Competition Structures via Weighted Mutual Information
author: | 
  | \large Zach Culp \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  |
  | \large Josie Peterburs \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  |
  | \large Ryan McShane \vspace{-1.1mm}
  | \normalsize University of Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  |
  | \large Gregory J. Matthews \vspace{-1.1mm}
  |	\normalsize  \vspace{-1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Center for Data Science and Consulting \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`email`](mailto:ypu@something.edu) \vspace{-1mm}
  | \normalsize [`zculp@luc.edu`](mailto:zculp@luc.edu) \vspace{-1mm}
  | \normalsize [`jpeterburs@luc.edu`](mailto:jpeterburs@luc.edu) \vspace{-1mm}
  | \normalsize [`rmcshane@uchicago.edu`](mailto:rmcshane@uchicago.edu) \vspace{-1mm}
  | \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1mm}
abstract: | 
 | We evaluate multiple competitions structures to assess their effectiveness in accurately reproducing the true underlying ranking of participating teams. Using repeated simulations for each structure, we compare teams’ true performance ranks to their simulated outcomes. To quantify the accuracy of each format, we calculate weighted mutual information between true and simulated rankings across a large number of replicates. "Salient" weights are applied to look at only the top n teams/competitors in a given competition structure. This information-theoretic approach allows us to objectively compare formats and identify which structures most effectively preserve rank order. Results are visualized through comparative plots, providing clear insights into the trade-offs and strengths of each competition design.
 | 
  \vspace{2mm}
  | Keywords: tournament structures, mutual information
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
editor_options: 
  markdown: 
    wrap: 72
---

@@

Here, we propose a novel metric for evaluating different competition structures using an information theoretic approach.  As an analogy, we view the results of a tournament as a transmitted message, and we use information theory to assess how accurately the message (i.e. the tournament results) matches the true transmission (i.e. the true ranking of competitors).

Specifically, we use weighted mutual information with a weights based on squared error loss^[other loss functions can easily be incorporated into our method] between the ranks generated by a competition structure and the true ranks of the competitors.  Additionally, we include an indicator vector to specify the salient ranks, allowing us to compare different competition structures with different choices for salient ranks.

The remainder of this manuscript presents the mathematical foundation of the proposed measure followed by a simulation study estimating out proposed measure for different types of competition structures with varying choices of salient ranks.  We close by offering a summary of the work and ideas for future directions.

@@

<!-- Different competition structures have unique strengths and weaknesses, affecting how well they reflect the true rankings of the teams. The goal of a tournament is to find the best teams, but how can we quantify how well a tournament performs? In some instances, the tournament organizers only award the overall winner, but other times, the top three or ten teams are awarded. Because of this, organizers may prefer one structure over another based on their needs. Tournament organizers must also take into account factors like cost, timeliness, entertainment value, and fairness when choosing a tournament structure. -->

<!-- To evaluate a tournament’s effectiveness, we propose a numerical metric that quantifies how accurately it orders teams based on their true rankings. Tournament results can be viewed as a "message" attempting to convey the true team rankings. However, various factors introduce noise, leading to information loss. We use principles from information theory to measure this information loss across multiple tournament simulations to assess the reliability of different formats. -->

@@

Consider a set of $n$ teams indexed from $i = 1, 2, \cdots, n$ each with an associated true strength parameter $\boldsymbol{\theta} = (\theta_1, \theta_2, \ldots, \theta_n)$ such that $(\theta_1 > \theta_2 > \ldots > \theta_n)$. Next, let
$r(\boldsymbol{\theta})$ be the indexes of $\boldsymbol{\theta}$ when $\theta$ is sorted from largest to smallest so that
$r(\boldsymbol{\theta}) = \{1, 2, \ldots, n-1, n\}$. Next define
$T(\phi, s)$ as a tournament with schedule $\phi$ and seeding structure $s$. $\phi$ contains all the information about the scheduling of teams, which could be fully known prior to the tournament (*e.g.* round robin) or determined as the tournament progresses (*e.g.* single elimination tournament). We then let $\hat{r}(T(\phi, s),\boldsymbol{\theta})$ be an $n$-dimensional vector-valued random variable that gives the results of a tournament as a vector of the indexes of the vector $\boldsymbol{\theta}$. The index in the first position of the vector $\hat{r}(T(\phi, s),\boldsymbol{\theta})$ indicates the index of the team that finished first, the index in the second position of the vector $\hat{r}(T(\phi, s),\boldsymbol{\theta})$ indicates the index of the team that finished second, and so on.

@@

If one views the true ranking of the teams in a tournament as a message to be sent to a receiver and the outcome of the tournament as a message that is received, we can measure the "goodness" of a tournament in terms of its ability to accurately transmit the true ranking. We can then use concepts from information theory to assess the ability of a tournament to correctly rank teams. Specifically, we start with the concept of mutual information (@Guiasu1977-nu). For two random variables $X$ and $Y$ mutual
information is defined as:

@@

In order to alleviate this problem, we instead consider *weighted*
mutual information (@Guiasu1977-nu). We want to give more weight to permutations from $\hat{r}$ that are "closer" to $r$.

In general, any loss function $l(r, \hat{r})$ can be used to define the weigting function:

@@

Here though, we choose to use squared error loss:

$$
w(r, \hat{r}) =
\begin{cases}
\frac{1}{\hat{r}'r}, & r \ne \hat{r} \\
1, & r = \hat{r}
\end{cases}
$$
<!-- Note that one is 1/N! -->

<!-- And this does work because consistently getting the order incorrect is perfect information. -->

However, because the number of teams that need to be accurately ordered differs based on the choice of salient ranks, a second weighting function, $m(r)$, is used to indicate which ranks are salient.  We call these the "salient weights" to differentiate from the other weighting function.  As an example, for a four-team tournament, if the organizer decides that only the winner is important, then $m(r) = \{1,0,0,0\}$ (i.e. essentially measuring effectivity).  While we only consider indicators (i.e. 0's or 1's) in the salient weights vector, there is no reason why you couldn't weight the ranks by relative levels of importance.  For instance, letting $m(r) = \{3,1,0,0\}$ indicates that the salient ranks are 1 and 2, but getting the true best competitor as the winner is more important than ranking the true second best competitor second.

With the addition of the weighting functions, the formula for mutual information becomes:

@@

Because this mutual information is unitless, the formula can be
standardized to be between 0 and 1 using the complement of Rajski's
distance (@Rajski1961): $$
1 - (1 - \frac{I(X,Y)}{\text{max\{}H(X), H(Y)\}})
$$
